# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
First let's read the data from the provided zip archive using _read.csv_ and _unz_ utility
```{r}
data <- read.csv(unz("activity.zip","activity.csv"),header=TRUE)
```
Here is a sneakpeak of the dataset that we are dealing with
```{r}
str(data)
```

Let's clean the dataset by ignoring NA values for now
```{r cleanUp}
tidyData <- data[complete.cases(data$steps),]
str(tidyData)
```
After we have tidy dataset let's try to answer some of the following questions

## What is mean total number of steps taken per day?

Here is frequency histogram of total steps taken 
```{r histogram}
totalSteps <- aggregate(tidyData$steps ~ tidyData$date,tidyData,sum)
colnames(totalSteps) <- c("date","totalSteps")
hist(totalSteps$totalSteps,col="red",xlab="Total Steps", main= "Number of Steps")
```
The above dataset has a mean of `r mean(totalSteps$totalSteps)` and median `r median(totalSteps$totalSteps)`

## What is the average daily activity pattern?

```{r plot}
totalPerInterval <- aggregate(tidyData$steps ~ tidyData$interval, tidyData, sum)
colnames(totalPerInterval) <- c("interval","totalSteps")
plot(totalPerInterval$interval,totalPerInterval$totalSteps,type="l",xlab="Interval", ylab="Total Steps")
```

The maximum number of steps throughout all days is observed during the following interval
```{r maxSteps interval}
totalPerInterval[which.max(totalPerInterval$totalSteps),]
```


## Imputing missing values
The above calculations omitted NAs for simplicity, but let's take a look at complete picture
```{r NAs in the set}
colSums(is.na(data))
```

Since there is a considerable amount of missing data, let's imput missing values with average of that interval throughout the days. We will utilize plyr library for this purpose
```{r imput}
require(plyr)
impute.mean <- function(x) replace(x,is.na(x),mean(x,na.rm=TRUE))
#new data set with imputed values
fullData <- ddply(data,~ interval,transform,steps = impute.mean(steps))
```

Here is a histogram of the total number of steps taken each day after imputing missing values
```{r histogram for updated set}
totalStepsFull <- aggregate(fullData$steps ~ fullData$date,fullData,sum)
colnames(totalStepsFull) <- c("date","totalSteps")
hist(totalStepsFull$totalSteps,col="red",xlab="Total Steps", main="Number of Steps")
```

The above dataset has a mean of `r mean(totalStepsFull$totalSteps)` and median `r median(totalStepsFull$totalSteps)`
We can see that the mean of the dataset after imputing missing values remained the same and median has shifted to match the mean which is expected since we replaced `NAs` with average 

## Are there differences in activity patterns between weekdays and weekends?

```{r weekend}
require(lattice)
fullData$date <- as.Date(fullData$date)
dayType <- weekdays(fullData$date)=="Sunday" | weekdays(fullData$date)=="Saturday"
fullData$dayType <- factor(dayType,labels = c("Weekend", "Weekday"))
totalStepsByDayType <- ddply(fullData,.(dayType,interval),summarize, totalSteps=sum(steps))
xyplot(totalSteps ~ interval | dayType, data= totalStepsByDayType, layout = c(1,2), type="l")
```

